---
title: "GBIF_point_plots"
author: "Abby Robinson"
date: "2025-03-19"
output: html_document
---
point plots showing total data spread from iNaturalist and samples sizes 

# data processing 
```{r load packages}
library(curl)
library(sf)
library(ggplot2)
library(lubridate)
library(spData)
library(ggmap)
library(sp)
library(adehabitatHR) ##kernal density package 
library(maps)
```

```{r load and filter data}
# use download key to load data from GBIF download 
data <- occ_download_get(key="0017454-250310093411724", overwrite=TRUE)
data <- occ_download_import(data)
butterfly_data <- as.data.frame(data)

length(butterfly_data$gbifID)

butterfly_data <- subset(butterfly_data, select = c(gbifID, lifeStage, eventDate, decimalLatitude, decimalLongitude, species) )
butterfly_adults <- subset(butterfly_data, lifeStage == "Adult")
length(butterfly_adults$gbifID)

butterfly_adults <- na.omit(butterfly_adults)
length(butterfly_adults$gbifID)

#rename columns
colnames(butterfly_adults)[1] ="id"
colnames(butterfly_adults)[2] ="life_stage"
colnames(butterfly_adults)[3] ="event_date"
colnames(butterfly_adults)[4] ="latitude"
colnames(butterfly_adults)[5] ="longitude"

head(butterfly_adults)

unique(butterfly_adults$species)

limenitis <- subset(butterfly_data, species == "Limenitis astyanax")
unique(butterfly_data$species)
unique(limenitis$infraspecificEpithet) #dataset only inlcudes limenitis arthemis astyanax morph 
```

```{r create julian day columns}
#separate date and time 
butterfly_adults[c('observed_on', 'observed_time')] <- str_split_fixed(butterfly_adults$event_date, 'T', 2)
butterfly_adults

# Adding dates to whole  dataset
dates <- parse_date_time(butterfly_adults$observed_on, "ymd")
months <- months(dates)
year <- year(dates)
day <- day(dates)
days_of_the_year <- yday(dates) #  days since Jan 1
butterfly_adults$julian_day <- days_of_the_year
butterfly_adults$year <- year
butterfly_adults$month <- months
butterfly_adults$day <- day
butterfly_adults


butterfly_adults
```

#Create latitudianl bins
```{r create lat bins}
#run analysis on all research grade records between 2014 - 2024 (10 years)
butterfly_adults <- subset(butterfly_adults, year %in% 2014:2024)
length(butterfly_adults$id)

# Next use the round() function to create a new column binning all latitudes into integer values.
butterfly_adults$latitudinal_bin <- round(butterfly_adults$latitude)
butterfly_adults <- butterfly_adults[order(butterfly_adults$latitudinal_bin, decreasing=F),]
butterfly_adults

#subset latitudinal bins of interest 
butterfly_adults_30_40 <- subset(butterfly_adults, latitudinal_bin %in% 30:40)

# Check to ensure the data was binned correctly.
head(butterfly_adults)
tail(butterfly_adults) 
length(butterfly_adults$id) 
unique(butterfly_adults$year)
```

#cap sightings by unique combinations of species, latitude, and date 
this helps correct for artificial inflation of sightings on specific dates (for example, if there was a butterfly surveying event that caused an abnormally high number of butterflies to be recorded in iNaturalist)

```{r cap data}
#creating grid system of approximate 7 km cells to correct for observer bias 
butterfly_adults$lat0_1 <- round(butterfly_adults$latitude, digits = 1)
butterfly_adults$long0_1 <- round(butterfly_adults$longitude, digits = 1)

butterfly_adults$coords0_1 <- paste(butterfly_adults$lat0_1, butterfly_adults$long0_1, sep= "-")

butterfly_adults$sp_coords_date_ID <- paste(butterfly_adults$species, butterfly_adults$coords0_1, butterfly_adults$observed_on, sep= ":")
length(butterfly_adults$id)
butterfly_adults_capped = butterfly_adults[!duplicated(butterfly_adults$sp_coords_date_ID),]
length(butterfly_adults_capped$id) 
#this removes duplicate sightings of the same species on the same day that are within 7km of each other 

unique(butterfly_adults_capped$species)
#subset by species 
battus_capped <- subset(butterfly_adults_capped, species == "Battus philenor")
astyanax_capped <- subset(butterfly_adults_capped, species == "Limenitis astyanax")
archippus_capped <- subset(butterfly_adults_capped, species == "Limenitis archippus")
plexippus_capped <- subset(butterfly_adults_capped, species == "Danaus plexippus")
```

# calculate sample size 
calculate total number of observations per species from iNaturalist within the latitudinal bounds of the study (western limit of -98, northern limit of 41, which corresponds to the point at which battus becomes very rare, according to previous research (Ries&Mullen2008))

```{r sample sizes}
setwd("/Users/abbyrobinson/Desktop/Phenology Publication/phenology_analysis/phenesse_csv_files") #set working directory 
estimates <- read.csv("GBIF_phenesse_estimates_df.csv") #load phenesse estimates data file from working directory 

#subset by species 
battus <- subset(estimates, species == "battus")
astyanax <- subset(estimates, species == "astyanax")
archippus <- subset(estimates, species == "archippus")
plexippus <- subset(estimates, species == "D_plexippus")

#total sample size 
sum(battus$sample_size)
sum(astyanax$sample_size)
sum(archippus$sample_size)
sum(plexippus$sample_size)

#mean sample size per latitudinal bin 
mean(battus$sample_size)
mean(astyanax$sample_size)
mean(archippus$sample_size)
mean(plexippus$sample_size)

#standard error in sample size per latitudinal bin
print(sd(battus$sample_size)/sqrt(length((battus$sample_size))))
print(sd(astyanax$sample_size)/sqrt(length((astyanax$sample_size))))
print(sd(archippus$sample_size)/sqrt(length((archippus$sample_size))))
print(sd(plexippus$sample_size)/sqrt(length((plexippus$sample_size))))     
```



# point plots 
```{r create sf options for plotting}

## point plots (not for kernel density)
bat1 <- st_as_sf(battus_capped, coords = c("longitude", "latitude"))
b1 <- ggplot() + geom_sf(data = bat1, color = "lightpink", alpha = 0.5) + theme_bw()

ast1 <- st_as_sf(astyanax_capped, coords = c("longitude", "latitude"))
as1 <- ggplot() + geom_sf(data = ast1, color = "royalblue", alpha = 0.5) + theme_bw()

arc1 <- st_as_sf(archippus_capped, coords = c("longitude", "latitude"))
ar1 <- ggplot() + geom_sf(data = arc1, color = "olivedrab", alpha = 0.5) + theme_bw()

plex1 <- st_as_sf(plexippus_capped, coords = c("longitude", "latitude"))
pl1 <- ggplot() + geom_sf(data = plex1, color = "orange", alpha = 0.5) + theme_bw()

#name world map for point plots below 
world <- map_data("world")
```

```{r battus point plot}
bat_pp <- ggplot() + 
  geom_polygon( data=world, aes(x=long, y=lat, group=group), color="grey", fill = "white", alpha=0.7) + 
  labs(x = "Longitude", y = "Latitude")  +
  geom_sf(data = bat1, color = "navyblue", alpha = 0.3) +
  geom_rect(aes(xmin = -98, xmax = -74, ymin = 29.5, ymax = 41.5),
            fill = "transparent", color = "black", size = 2) +
  coord_sf(xlim = c(-130, -60), ylim = c(15, 60), expand = FALSE) +
  annotate("text", x=-129, y=59, label= "A.", fontface = "bold", size = 6, hjust = 0) + 
  annotate("text", x=-125, y=59, label= "B. philenor", fontface = "italic", size = 6, hjust = 0) + 
  annotate("text", x=-125, y=56, label= "n = 13650", size = 6, hjust = 0) + 
  theme_classic() + 
  theme(axis.ticks = element_line(size = 1, color="black") , 
        axis.ticks.length = unit(.3, "cm"), 
        axis.title = element_text( color="black", size=15, face=2),
        axis.line = element_line(size = 1, colour = "black", linetype=1),
        axis.text = element_text(color="black", size=15, face=1)
  ) 
bat_pp
```

```{r astyanax point plot}
ast_pp <- ggplot() + 
  geom_polygon( data=world, aes(x=long, y=lat, group=group), color="grey", fill = "white", alpha=0.7) + 
  labs(x = "Longitude", y = "Latitude")  +
  geom_sf(data = ast1, color = "lightblue", alpha = 0.3) +
  geom_rect(aes(xmin = -98, xmax = -74, ymin = 29.5, ymax = 41.5),
            fill = "transparent", color = "black", size = 2) +
  coord_sf(xlim = c(-130, -60), ylim = c(15, 60), expand = FALSE) +
  annotate("text", x=-129, y=59, label= "B.", fontface = "bold", size = 6, hjust = 0) + 
  annotate("text", x=-125, y=59, label= "L. a. astyanax", fontface = "italic", size = 6, hjust = 0) + 
  annotate("text", x=-125, y=56, label= "n = 13796", size = 6, hjust = 0) + 
  theme_classic() + 
  theme(axis.ticks = element_line(size = 1, color="black") , 
        axis.ticks.length = unit(.3, "cm"), 
        axis.title = element_text( color="black", size=15, face=2),
        axis.line = element_line(size = 1, colour = "black", linetype=1),
        axis.text = element_text(color="black", size=15, face=1)
  ) 
ast_pp
```

```{r}
arc_pp <- ggplot() + 
  geom_polygon( data=world, aes(x=long, y=lat, group=group), color="grey", fill = "white", alpha=0.7) + 
  labs(x = "Longitude", y = "Latitude")  +
  geom_sf(data = arc1, color = "orange", alpha = 0.3) +
  geom_rect(aes(xmin = -98, xmax = -74, ymin = 29.5, ymax = 41.5),
            fill = "transparent", color = "black", size = 2) +
  coord_sf(xlim = c(-130, -60), ylim = c(15, 60), expand = FALSE) +
  annotate("text", x=-129, y=59, label= "D.", fontface = "bold", size = 6, hjust = 0) + 
  annotate("text", x=-125, y=59, label= "L. archippus", fontface = "italic", size = 6, hjust = 0) + 
  annotate("text", x=-125, y=56, label= "n = 7062", size = 6, hjust = 0) + 
  theme_classic() + 
  theme(axis.ticks = element_line(size = 1, color="black") , 
        axis.ticks.length = unit(.3, "cm"), 
        axis.title = element_text( color="black", size=15, face=2),
        axis.line = element_line(size = 1, colour = "black", linetype=1),
        axis.text = element_text(color="black", size=15, face=1)
  ) 
arc_pp
```

```{r}
plex_pp <- ggplot() + 
  geom_polygon( data=world, aes(x=long, y=lat, group=group), color="grey", fill = "white", alpha=0.7) + 
  labs(x = "Longitude", y = "Latitude")  +
  geom_sf(data = plex1, color = "brown", alpha = 0.3) +
  geom_rect(aes(xmin = -98, xmax = -74, ymin = 29.5, ymax = 41.5),
            fill = "transparent", color = "black", size = 2) +
  coord_sf(xlim = c(-130, -60), ylim = c(15, 60), expand = FALSE) +
  annotate("text", x=-129, y=59, label= "C.", fontface = "bold", size = 6, hjust = 0) + 
  annotate("text", x=-125, y=59, label= "D. plexippus", fontface = "italic", size = 6, hjust = 0) + 
  annotate("text", x=-125, y=56, label= "n = 39636", size = 6, hjust = 0) + 
  theme_classic() + 
  theme(axis.ticks = element_line(size = 1, color="black") , 
        axis.ticks.length = unit(.3, "cm"), 
        axis.title = element_text( color="black", size=15, face=2),
        axis.line = element_line(size = 1, colour = "black", linetype=1),
        axis.text = element_text(color="black", size=15, face=1)
  ) 
plex_pp
```


```{r save figure}

library(patchwork)

final_fig <- bat_pp + plex_pp + ast_pp + arc_pp 

ggsave(final_fig, file = "/Users/abbyrobinson/Desktop/Phenology Publication/phenology_analysis/phenology_figures/final_point_plot.pdf", width=15, height=10, units=c("in"), useDingbats=FALSE)
```