---
title: "phenesse_analysis_GBIF"
author: "Abby Robinson"
date: "2025-03-19"
output: html_document
---

this script includes all analyses of phenesee estimates 

# Calculating a p-value from confidence intervals 
Code Source: https://genometoolbox.blogspot.com/2013/11/how-to-estimate-p-value-from-confidence.html#:~:text=Therefore%2C%20if%20we%20are%20given,to%20calculate%20the%20p%2Dvalue.

All code used to calculate z-scores and p-values were adapted from the above source.

# data processing 
```{r load data and subset species}
setwd("/Users/abbyrobinson/Desktop/Phenology Publication/phenology_analysis/phenesse_csv_files") # set working directory 
estimates <- read.csv("GBIF_phenesse_estimates_df.csv") # load phenesse estimate data 

#remove NAs from dataset 
length(estimates$species) 

## we want to calculate z-scores for two comparisons. subset out those species before proceeding 

battus <- subset(estimates, species == "battus")
astyanax <- subset(estimates, species == "astyanax")

plexippus <- subset(estimates, species == "D_plexippus")
archippus <- subset(estimates, species == "archippus")

```

#calculate standard deviation for each species

```{r battus}

##check data structure and convert latitude to factor and sample size into integer data 
str(battus)
battus$latitude = as.factor(battus$latitude)
battus$sample_size = as.numeric(battus$sample_size)
battus$offset_high_ci = as.numeric(battus$offset_high_ci)
str(battus)

#split data by latitudinal bin for loop function 
my_battus_splits = split(battus, battus$latitude)
my_battus_splits$"30"$latitude
length(my_battus_splits) #11 latitudes 
as.vector(unique(my_battus_splits$"30"$latitude))

sd_battus_df = data.frame() # create empty data frame 

for (i in my_battus_splits) {
  ci_diff <- (i$onset_high_ci - i$onset_low_ci)/2
  se <- ci_diff/1.96  #1.96 is the number corresponding to the 95% confidence interval 
  sd <- se*sqrt(i$sample_size)
  lat <- as.vector(unique(i$latitude))
  output = matrix(
        c(lat, sd), 
        nrow = 1,             
        ncol = 2,           
        byrow = TRUE         
        )
  output <- as.data.frame(output)
  sd_battus_df = rbind(sd_battus_df, output)
}

sd_battus_df # data frame with standard error for battus 

##check loop by running lat 30 by hand  
battus30 <- subset(battus, latitude == 30)
ci_diff <- (battus30$onset_high_ci - battus30$onset_low_ci)/2
se <- ci_diff/1.96
sd <- se*sqrt(battus30$sample_size)
sd #matches lat 30 in loop 

```

```{r astyanax }

##check data structure and convert latitude to factor and sample size into integer data 
str(astyanax)
astyanax$latitude = as.factor(astyanax$latitude)
astyanax$sample_size = as.numeric(astyanax$sample_size)
astyanax$offset_high_ci = as.numeric(astyanax$offset_high_ci)
str(astyanax)

#split data by latitudinal bin for loop function 
my_astyanax_splits = split(astyanax, astyanax$latitude)

sd_astyanax_df = data.frame() 

for (i in my_astyanax_splits) {
  ci_diff <- (i$onset_high_ci - i$onset_low_ci)/2
  se <- ci_diff/1.96  #1.96 is the number corresponding to the 95% confidence interval 
  sd <- se*sqrt(i$sample_size)
  lat <- as.vector(unique(i$latitude))
  output = matrix(
        c(lat, sd), 
        nrow = 1,             
        ncol = 2,           
        byrow = TRUE         
        )
  output <- as.data.frame(output)
  sd_astyanax_df = rbind(sd_astyanax_df, output)
}

sd_astyanax_df

##check loop by running lat 30 by hand  
astyanax30 <- subset(astyanax, latitude == 30)
ci_diff <- (astyanax30$onset_high_ci - astyanax30$onset_low_ci)/2
se <- ci_diff/1.96
sd <- se*sqrt(astyanax30$sample_size)
sd #matches result from loop function 
```

```{r archippus }

##check data structure and convert latitude to factor and sample size into integer data 
str(archippus)
archippus$latitude = as.factor(archippus$latitude)
archippus$sample_size = as.numeric(archippus$sample_size)
str(archippus)

my_archippus_splits = split(archippus, archippus$latitude)

sd_archippus_df = data.frame()

for (i in my_archippus_splits) {
  ci_diff <- (i$onset_high_ci - i$onset_low_ci)/2
  se <- ci_diff/1.96  #1.96 is the number corresponding to the 95% confidence interval 
  sd <- se*sqrt(i$sample_size)
  lat <- as.vector(unique(i$latitude))
  output = matrix(
        c(lat, sd), 
        nrow = 1,             
        ncol = 2,           
        byrow = TRUE         
        )
  output <- as.data.frame(output)
  sd_archippus_df = rbind(sd_archippus_df, output)
}

sd_archippus_df

```

```{r plexippus }

##check data structure and convert latitude to factor and sample size into integer data 
str(plexippus)
plexippus$latitude = as.factor(plexippus$latitude)
plexippus$sample_size = as.numeric(plexippus$sample_size)
plexippus$offset_high_ci = as.numeric(plexippus$offset_high_ci)
str(plexippus)

#split data by latitudinal bin for loop function 
my_plexippus_splits = split(plexippus, plexippus$latitude)

sd_plexippus_df = data.frame() 

for (i in my_plexippus_splits) {
  ci_diff <- (i$onset_high_ci - i$onset_low_ci)/2
  se <- ci_diff/1.96  #1.96 is the number corresponding to the 95% confidence interval 
  sd <- se*sqrt(i$sample_size)
  lat <- as.vector(unique(i$latitude))
  output = matrix(
        c(lat, sd), 
        nrow = 1,             
        ncol = 2,           
        byrow = TRUE         
        )
  output <- as.data.frame(output)
  sd_plexippus_df = rbind(sd_plexippus_df, output)
}

sd_plexippus_df

```

# create data frames for z tests
```{r create data frame}
##generate data frame with latitude, estimate_battus, estimate_astyanax, n_bat, n_ast, sd_bat, sd_ast 

ztest_data_df <- data.frame(latitude = 30:40, 
                             battus_estimate = battus$onset_estimate, 
                             battus_n = battus$sample_size, 
                             battus_sd = sd_battus_df$V2, 
                             astyanax_estimate = astyanax$onset_estimate, 
                             astyanax_n = astyanax$sample_size, 
                             astyanax_sd = sd_astyanax_df$V2,
                             plexippus_estimate = plexippus$onset_estimate, 
                             plexippus_n = plexippus$sample_size, 
                             plexippus_sd = sd_plexippus_df$V2, 
                             archippus_estimate = archippus$onset_estimate, 
                             archippus_n = archippus$sample_size, 
                             archippus_sd = sd_archippus_df$V2)
ztest_data_df

```
#run z tests 
```{r z test loop function}
## write loop function that will run the z test and p value calculation for each pairwise comparison 

ztest_data_df

##check data structure and convert latitude to factor and sample size into integer data 
str(ztest_data_df)

ztest_data_df$latitude = as.factor(ztest_data_df$latitude)
ztest_data_df$battus_sd = as.numeric(ztest_data_df$battus_sd)
ztest_data_df$astyanax_sd = as.numeric(ztest_data_df$astyanax_sd)
ztest_data_df$plexippus_sd = as.numeric(ztest_data_df$plexippus_sd)
ztest_data_df$archippus_sd = as.numeric(ztest_data_df$archippus_sd)
str(ztest_data_df)

#split z score data frame by latitude for loop function
my_ztest_splits = split(ztest_data_df, ztest_data_df$latitude)

options(scipen=999)

ztest_df = data.frame() #create empty data frame 

##this loop function runs a two sample z test for all pairwise comparisons at each latitude 
for (i in my_ztest_splits) {
  ## battus - astyanax comparison 
  bat_ast_numerator= (i$battus_estimate - i$astyanax_estimate) - 0  #because we assume that the two treatments have the same effect, we then have to assume that the difference between the population means is zero.
  bat_ast_denominator= sqrt(((i$battus_sd^2)/i$battus_n)+((i$astyanax_sd^2)/i$astyanax_n))
  bat_ast_z_score = abs(bat_ast_numerator/bat_ast_denominator)
  bat_ast_p_value = 2*pnorm(q=bat_ast_z_score, lower.tail=FALSE) #q is the z score 
  lat = as.vector(unique(i$latitude))
  bat_ast = matrix(
        c(lat, "battus_astyanax", bat_ast_z_score, bat_ast_p_value), 
        nrow = 1,             
        ncol = 4,           
        byrow = TRUE         
        )
  bat_ast <- as.data.frame(bat_ast)
  
  ## plexippus - archippus comparison 
  plex_arc_numerator= (i$plexippus_estimate - i$archippus_estimate) - 0  #because we assume that the two treatments have the same effect, we then have to assume that the difference between the population means is zero.
  plex_arc_denominator= sqrt(((i$plexippus_sd^2)/i$plexippus_n)+((i$archippus_sd^2)/i$archippus_n))
  plex_arc_z_score = abs(plex_arc_numerator/plex_arc_denominator)
  plex_arc_p_value = 2*pnorm(q=plex_arc_z_score, lower.tail=FALSE) #q is the z score 
  lat = as.vector(unique(i$latitude))
  plex_arc = matrix(
        c(lat, "plexippus_archippus", plex_arc_z_score, plex_arc_p_value), 
        nrow = 1,             
        ncol = 4,           
        byrow = TRUE         
        )
  plex_arc <- as.data.frame(plex_arc)
  
  ztest_df = rbind(ztest_df, bat_ast, plex_arc)
}

colnames(ztest_df) <- c("latitude", "species_comparison", "z_score", "p_value") 
ztest_df


## write output into a CSV file 
write.csv(ztest_df,file='/Users/abbyrobinson/Desktop/Phenology Publication/phenology_analysis/phenesse_csv_files/ztest_df.csv', row.names=FALSE)

## pvalue correction
ztest_bat_ast <- subset(ztest_df, species_comparison == "battus_astyanax")
ztest_plex_arc <- subset(ztest_df, species_comparison == "plexippus_archippus")

#battus - astyanax comparison
p_value_holm_adjusted <- p.adjust(ztest_bat_ast$p_value, method = "holm", n = length(ztest_bat_ast$p_value)) 
#n= number of comparisons, which is 11 for each species pair 

p_value_holm_adjusted <- round(p_value_holm_adjusted, digits = 5) 
ztest_bat_ast$p_value_holm_adjusted <- p_value_holm_adjusted

#plexippus - archippus comparison
p_value_holm_adjusted <- p.adjust(ztest_plex_arc$p_value, method = "holm", n = length(ztest_plex_arc$p_value)) 
p_value_holm_adjusted <- round(p_value_holm_adjusted, digits = 5) 

ztest_plex_arc$p_value_holm_adjusted <- p_value_holm_adjusted

ztest_p_adjusted <- rbind(ztest_bat_ast, ztest_plex_arc)

## write output into a CSV file 
write.csv(ztest_p_adjusted,file='/Users/abbyrobinson/Desktop/Phenology Publication/phenology_analysis/phenesse_csv_files/ztest_p_adjusted.csv', row.names=FALSE)
```

#calculate average emergence delay between models and mimics 

```{r mean emergence delay}
## emergence delay at each latitude with standard error

#battus-astyanas comparison
ztest_bat_ast <- subset(ztest_df, species_comparison == "battus_astyanax")
ztest_data_df # data file with onset estimates 

ztest_bat_ast$battus_estimate <- ztest_data_df$battus_estimate
ztest_bat_ast$astyanax_estimate <- ztest_data_df$astyanax_estimate
emergence_delay <- ztest_bat_ast$astyanax_estimate - ztest_bat_ast$battus_estimate 

ztest_bat_ast$emergence_delay <- emergence_delay


mean(ztest_bat_ast$emergence_delay) #15.75455 days 
# mean emergence delay between models and mimics 

#standard error
print(sd(ztest_bat_ast$emergence_delay)/sqrt(length((ztest_bat_ast$emergence_delay))))
#1.785961

## plexippus-archippus comparison
ztest_plex_arc <- subset(ztest_df, species_comparison == "plexippus_archippus")
ztest_data_df # data file with onset estimates 

ztest_plex_arc$plexippus_estimate <- ztest_data_df$plexippus_estimate
ztest_plex_arc$archippus_estimate <- ztest_data_df$archippus_estimate
emergence_delay2 <- ztest_plex_arc$archippus_estimate - ztest_plex_arc$plexippus_estimate 

ztest_plex_arc$emergence_delay <- emergence_delay2

mean(ztest_plex_arc$emergence_delay) #9.818182 days

#standard error
print(sd(ztest_plex_arc$emergence_delay)/sqrt(length((ztest_plex_arc$emergence_delay))))
#5.962196

##mean emergence delay between plexippus and archippus only in latitudes where plexippus emerged first 
ztest_plex_arc_Mod_first <- subset(ztest_plex_arc, latitude %in% 30:37)
mean(ztest_plex_arc_Mod_first$emergence_delay) #20 days 

#standard error
print(sd(ztest_plex_arc_Mod_first$emergence_delay)/sqrt(length((ztest_plex_arc_Mod_first$emergence_delay)))) #3.455379

##mean emergence delay between plexippus and archippus only in latitudes where archippus emerged first  
ztest_plex_arc_Mim_first <- subset(ztest_plex_arc, latitude %in% 38:40)
mean(ztest_plex_arc_Mim_first$emergence_delay) #17.33333 days 

#standard error
print(sd(ztest_plex_arc_Mim_first$emergence_delay)/sqrt(length((ztest_plex_arc_Mim_first$emergence_delay)))) #5.783117

## write output into a CSV file 
write.csv(ztest_bat_ast,file='/Users/abbyrobinson/Desktop/Phenology Publication/phenology_analysis/phenesse_csv_files/ztest_bat_ast.csv', row.names=FALSE)

## write output into a CSV file 
write.csv(ztest_plex_arc,file='/Users/abbyrobinson/Desktop/Phenology Publication/phenology_analysis/phenesse_csv_files/ztest_plex_arc.csv', row.names=FALSE)

```
# mean sample size & standard error for each species 
```{r mean sample size}
## battus 

#mean 
battus_n <- ztest_data_df$battus_n
battus_mean_n <- mean(battus_n)
#standard error 
print(sd(battus_n)/sqrt(length((battus_n))))

## astyanax 

#mean 
astyanax_n <- ztest_data_df$astyanax_n
astyanax_mean_n <- mean(astyanax_n)
#standard error 
print(sd(astyanax_n)/sqrt(length((astyanax_n))))

## plexippus

#mean 
plexippus_n <- ztest_data_df$plexippus_n
plexippus_mean_n <- mean(plexippus_n)
#standard error 
print(sd(plexippus_n)/sqrt(length((plexippus_n))))


## archippus

#mean 
archippus_n <- ztest_data_df$archippus_n
archippus_mean_n <- mean(archippus_n)
#standard error 
print(sd(archippus_n)/sqrt(length((archippus_n))))


```

# mean onset dates for each species 
```{r mean onset date}

##subset out individual species 
battus <- subset(estimates, species == "battus")
astyanax <- subset(estimates, species == "astyanax")
plexippus <- subset(estimates, species == "D_plexippus")
archippus <- subset(estimates, species == "archippus")

mean(battus$onset_estimate) #97.85
battus_onset <- battus$onset_estimate
#standard error 
print(sd(battus_onset)/sqrt(length((battus_onset)))) #5.351436

mean(astyanax$onset_estimate) #113.6045
astyanax_onset <- astyanax$onset_estimate
print(sd(astyanax_onset)/sqrt(length((astyanax_onset)))) #6.455162

mean(plexippus$onset_estimate) #113.3091
plexippus_onset <- plexippus$onset_estimate
print(sd(plexippus_onset)/sqrt(length((plexippus_onset)))) #11.31648

mean(archippus$onset_estimate) #123.1273
archippus_onset <- archippus$onset_estimate
print(sd(archippus_onset)/sqrt(length((archippus_onset)))) #5.446268

#difference in onset date between northernmost and southernmost latitudes 
battus_north <- subset(battus, latitude == 40)
battus_north$onset_estimate
battus_south <- subset(battus, latitude == 30)
battus_south$onset_estimate
battus_onset_diff <- battus_north$onset_estimate - battus_south$onset_estimate

astyanax_north <- subset(astyanax, latitude == 40)
astyanax_north$onset_estimate
astyanax_south <- subset(astyanax, latitude == 30)
astyanax_south$onset_estimate
astyanax_onset_diff <- astyanax_north$onset_estimate - astyanax_south$onset_estimate

plexippus_north <- subset(plexippus, latitude == 40)
plexippus_north$onset_estimate
plexippus_south <- subset(plexippus, latitude == 30)
plexippus_south$onset_estimate
plexippus_onset_diff <- plexippus_north$onset_estimate - plexippus_south$onset_estimate

archippus_north <- subset(archippus, latitude == 40)
archippus_north$onset_estimate
archippus_south <- subset(archippus, latitude == 30)
archippus_south$onset_estimate
archippus_onset_diff <- archippus_north$onset_estimate - archippus_south$onset_estimate

onset_diff <- c(battus_onset_diff, astyanax_onset_diff, plexippus_onset_diff, archippus_onset_diff)
mean(onset_diff) #71.5
print(sd(onset_diff)/sqrt(length((onset_diff)))) #15.30523

```