---
title: "phenesse_estimates"
author: "Abby Robinson"
date: "2024-01-23"
output: html_document
---

This code document loads data files downloaded from iNaturalist, filters out subspecies and data entries with missing information, etc., organizes data for phenology analysis, and runs phenesse quantile functions to generate file with phenology estimates. Only observations with photos that were classified as research grade were included in this analysis 

#Data Processing 

```{r load packages }
library(curl)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(lubridate)
library(sf)
library(spData)
library(phenesse)
```


```{r B. philenor data}
setwd("/Users/abbyrobinson/Desktop/Phenology Publication/Phenology_R_Scripts") # set working directory 
battus <- read.csv("b_philenor_observations_1-23-2024.csv") # read data file from iNaturalist 
head(battus)

length(battus$id) 
b <- subset(battus, select = c(id, observed_on, latitude, longitude, scientific_name) )
head(b)

b <- b %>% filter(scientific_name %in% c("Battus philenor", "Battus philenor philenor")) #getting rid of other subspecies, e.g. B. p. hirsuta
length(b$id) 
battus <- na.omit(b)
length(battus$id) 

# Adding dates to whole Battus dataset
b.dates <- parse_date_time(battus$observed_on, "ymd")
b.months <- months(b.dates)
b.year <- year(b.dates)
b.day <- day(b.dates)
days_of_the_year <- yday(b.dates) #  days since Jan 1
battus$julian_day <- days_of_the_year
battus$year <- b.year
battus$month <- b.months
battus$day <- b.day
battus

# Adding longitudinal cutoff (see rationale for this cutoff in the plots made later in this document & in phenology point plot figure)
battus <- battus %>% filter(longitude >= -98)
#summary(battus)
battus
```

```{r L. a. astyanax data}
setwd("/Users/abbyrobinson/Desktop/Phenology Publication/Phenology_R_Scripts")
astyanax <- read.csv("L.a.astyanax_observations_1-23-24.csv")
head(astyanax)

length(astyanax$id) 
l <- subset(astyanax, select = c(id, observed_on, latitude, longitude, scientific_name) )

l <- l %>% filter(scientific_name %in% "Limenitis arthemis astyanax") # getting rid of other subspecies (namely L. a. arizonensis)
length(l$id) 
astyanax <- na.omit(l)
length(astyanax$id) 

# Adding dates to whole Limenitis dataset
l.dates <- parse_date_time(astyanax$observed_on, "ymd")
l.months <- months(l.dates)
l.year <- year(l.dates)
l.day <- day(l.dates)
days_of_the_year <- yday(l.dates) #  days since Jan 1
astyanax$julian_day <- days_of_the_year
astyanax$year <- l.year
astyanax$month <- l.months
astyanax$day <- l.day
astyanax

# Adding longitudinal cutoff
astyanax <- astyanax %>% filter(longitude >= -98)
#summary(astyanax)
astyanax
```

```{r L. archippus data}
setwd("/Users/abbyrobinson/Desktop/Phenology Publication/Phenology_R_Scripts")
archippus <- read.csv("L.archippus_observations_1-23-2024.csv")
head(archippus)

length(archippus$id)
a <- subset(archippus, select = c(id, observed_on, latitude, longitude, scientific_name) )
head(a)

archippus <- na.omit(a)
length(archippus$id)

# Adding dates to whole archippus dataset
a.dates <- parse_date_time(archippus$observed_on, "ymd")
a.months <- months(a.dates)
a.year <- year(a.dates)
a.day <- day(a.dates)
days_of_the_year <- yday(a.dates) #  days since Jan 1
archippus$julian_day <- days_of_the_year
archippus$year <- a.year
archippus$month <- a.months
archippus$day <- a.day
archippus

# Adding longitudinal cutoff
archippus <- archippus %>% filter(longitude >= -98)
#summary(archippus)
archippus
```
#Create latitudianl bins
```{r B. philenor}
# we are only using data from 2019-2023, when iNaturalist data density increased significantly 
battus <- subset(battus, year %in% 2019:2023)


# Next use the round() function to create a new column binning all latitudes into integer values.
battus$latitudinal_bin <- round(battus$latitude)
battus <- battus[order(battus$latitudinal_bin, decreasing=F),]
battus

# Check to ensure the data was binned correctly.
head(battus)
tail(battus) 
length(battus$id) 
unique(battus$year)
```

```{r L. a. astyanax}
# First subset all data by the total latitudinal range of interest.
astyanax <- subset(astyanax, year %in% 2019:2023)

# Next use the round() function to create a new column binning all latitudes into integer values.
astyanax$latitudinal_bin <- round(astyanax$latitude)
astyanax <- astyanax[order(astyanax$latitudinal_bin, decreasing=F),]
astyanax

# Check to ensure the data was binned correctly.
head(astyanax)
tail(astyanax) 
length(astyanax$id) 
unique(astyanax$year)
```


```{r L. archippus }
# First subset all data by the total latitudinal range of interest.
archippus  <- subset(archippus, year %in% 2019:2023)

# Next use the round() function to create a new column binning all latitudes into integer values.
archippus$latitudinal_bin <- round(archippus$latitude)
archippus <- archippus[order(archippus$latitudinal_bin, decreasing=F),]
archippus

# Check to ensure the data was binned correctly.
head(archippus)
tail(archippus) 
length(archippus$id)
unique(archippus$year)
```

#cap sightings by latitude and date 
this helps correct for artificial inflation of sightings on specific dates (for example, if there was a butterfly surveying event that caused an abnormally high number of butterflies to be recorded in iNaturalist)

```{r all species}
battus$lat_obs <- paste(battus$latitudinal_bin, battus$observed_on, sep= ":")
astyanax$lat_obs <- paste(astyanax$latitudinal_bin, astyanax$observed_on, sep= ":")
archippus$lat_obs <- paste(archippus$latitudinal_bin, archippus$observed_on, sep= ":")

battus_capped = battus[!duplicated(battus$lat_obs),]
astyanax_capped = astyanax[!duplicated(astyanax$lat_obs),]
archippus_capped = archippus[!duplicated(archippus$lat_obs),]
```

# check data structure 
```{r checked with archippus}
# checking the structure of our data and what type of data you have
str(archippus_capped)
archippus_capped$latitudinal_bin = as.factor(archippus_capped$latitudinal_bin)
```

#Phenesse estimates 

```{r archippus estimates}
#archippus estimates 
my_arc_splits = split(archippus_capped, archippus_capped$latitudinal_bin)
my_arc_splits$"30"$julian_day
latitude <- as.vector(unique(my_arc_splits$"30"$latitudinal_bin))

archippus_df = data.frame()

# archippus 
# for loop with quantile estimates 
for (i in my_arc_splits) {
  sample_size <- length(i$id)
  if (length(i$id) > 30){
    onset <- quantile_ci(observations = i$julian_day, percentile = 0.05, bootstraps = 10000, conf = 0.95, type = "perc")
    offset <- quantile_ci(observations = i$julian_day, percentile = 0.95, bootstraps = 10000, conf = 0.95, type = "perc")
    output <- c("archippus", as.vector(unique(i$latitudinal_bin)), sample_size, onset, offset)
    output <- as.data.frame(output)
    colnames(output) <- c("species", "latitude", "sample_size", "onset_estimate", "onset_low_ci", "onset_high_ci", "offset_estimate", "offset_low_ci", "offset_high_ci") 
    } else {
      output = matrix(
        c("archippus", as.vector(unique(i$latitudinal_bin)), sample_size, "NA", "NA", "NA", "NA", "NA", "NA"), 
        nrow = 1,             
        ncol = 9,           
        byrow = TRUE         
        )
      output <- as.data.frame(output)
      colnames(output) <- c("species", "latitude", "sample_size", "onset_estimate", "onset_low_ci", "onset_high_ci", "offset_estimate", "offset_low_ci", "offset_high_ci") 
    }

  archippus_df = rbind(archippus_df, output)
  
}
archippus_df
```

```{r astyanax estimates}
# astyanax estimates 
my_ast_splits = split(astyanax_capped, astyanax_capped$latitudinal_bin)
astyanax_df = data.frame()

# for loop with quantile estimates 
for (i in my_ast_splits) {
  sample_size <- length(i$id)
  if (length(i$id) > 30){
    onset <- quantile_ci(observations = i$julian_day, percentile = 0.05, bootstraps = 10000, conf = 0.95, type = "perc")
    offset <- quantile_ci(observations = i$julian_day, percentile = 0.95, bootstraps = 10000, conf = 0.95, type = "perc")
    output <- c("astyanax", as.vector(unique(i$latitudinal_bin)), sample_size, onset, offset)
    output <- as.data.frame(output)
    colnames(output) <- c("species", "latitude", "sample_size", "onset_estimate", "onset_low_ci", "onset_high_ci", "offset_estimate", "offset_low_ci", "offset_high_ci") 
    } else {
      output = matrix(
        c("astyanax", as.vector(unique(i$latitudinal_bin)), sample_size, "NA", "NA", "NA", "NA", "NA", "NA"), 
        nrow = 1,             
        ncol = 9,           
        byrow = TRUE         
        )
      output <- as.data.frame(output)
      colnames(output) <- c("species", "latitude", "sample_size", "onset_estimate", "onset_low_ci", "onset_high_ci", "offset_estimate", "offset_low_ci", "offset_high_ci") 
    }

  astyanax_df = rbind(astyanax_df, output)
  
}

astyanax_df
```


```{r battus estimates }
# battus estimates 
my_bat_splits = split(battus_capped, battus_capped$latitudinal_bin)
battus_df = data.frame()

# for loop with quantile estimates 
for (i in my_bat_splits) {
  sample_size <- length(i$id)
  if (length(i$id) > 30){
    onset <- quantile_ci(observations = i$julian_day, percentile = 0.05, bootstraps = 10000, conf = 0.95, type = "perc")
    offset <- quantile_ci(observations = i$julian_day, percentile = 0.95, bootstraps = 10000, conf = 0.95, type = "perc")
    output <- c("battus", as.vector(unique(i$latitudinal_bin)), sample_size, onset, offset)
    output <- as.data.frame(output)
    colnames(output) <- c("species", "latitude", "sample_size", "onset_estimate", "onset_low_ci", "onset_high_ci", "offset_estimate", "offset_low_ci", "offset_high_ci") 
    } else {
      output = matrix(
        c("battus", as.vector(unique(i$latitudinal_bin)), sample_size, "NA", "NA", "NA", "NA", "NA", "NA"), 
        nrow = 1,             
        ncol = 9,           
        byrow = TRUE         
        )
      output <- as.data.frame(output)
      colnames(output) <- c("species", "latitude", "sample_size", "onset_estimate", "onset_low_ci", "onset_high_ci", "offset_estimate", "offset_low_ci", "offset_high_ci") 
    }

  battus_df = rbind(battus_df, output)
  
}

battus_df
```

#create data file and save externally 
```{r write csv file}
phenesse_estimates_df = rbind(archippus_df, astyanax_df, battus_df)

## write output into a CSV file 
write.csv(phenesse_estimates_df,file='/Users/abbyrobinson/Desktop/Phenology Publication/phenology_analysis/phenesse_estimates_df.csv', row.names=FALSE)

```


