---
title: "point_plots"
author: "Abby Robinson"
date: "2024-01-25"
output: html_document
---

point plots showing total data spread from iNaturalist and samples sizes 

# data processing 
```{r load packages}
library(curl)
library(sf)
library(ggplot2)
library(lubridate)
library(spData)
library(ggmap)
library(sp)
library(adehabitatHR) ##kernal density package 
library(maps)
```

```{r B. philenor data}
setwd("/Users/abbyrobinson/Desktop/Phenology Publication/Phenology_R_Scripts") # set working directory
battus <- read.csv("b_philenor_observations_1-23-2024.csv") # load inat data file 
head(battus)

length(battus$id) 
b <- subset(battus, select = c(id, observed_on, latitude, longitude, scientific_name) )
head(b)

b <- subset(b, scientific_name %in% c("Battus philenor", "Battus philenor philenor") )
length(b$id) 
battus <- na.omit(b)
length(battus$id) 

# Adding dates to whole Battus dataset
b.dates <- parse_date_time(battus$observed_on, "ymd")
b.months <- months(b.dates)
b.year <- year(b.dates)
b.day <- day(b.dates)
days_of_the_year <- yday(b.dates) #  days since Jan 1
battus$julian_day <- days_of_the_year
battus$year <- b.year
battus$month <- b.months
battus$day <- b.day
battus

#summary(battus)
battus
```

```{r L. a. astyanax data}
setwd("/Users/abbyrobinson/Desktop/Phenology Publication/Phenology_R_Scripts")
astyanax <- read.csv("L.a.astyanax_observations_1-23-24.csv")
head(astyanax)

length(astyanax$id) 
l <- subset(astyanax, select = c(id, observed_on, latitude, longitude, scientific_name) )


l <- subset(l, scientific_name %in% c("Limenitis arthemis astyanax") )
length(l$id) 
astyanax <- na.omit(l)
length(astyanax$id) 

# Adding dates to whole Limenitis dataset
l.dates <- parse_date_time(astyanax$observed_on, "ymd")
l.months <- months(l.dates)
l.year <- year(l.dates)
l.day <- day(l.dates)
days_of_the_year <- yday(l.dates) #  days since Jan 1
astyanax$julian_day <- days_of_the_year
astyanax$year <- l.year
astyanax$month <- l.months
astyanax$day <- l.day
astyanax

#summary(astyanax)
astyanax
```

```{r L. archippus data}
setwd("/Users/abbyrobinson/Desktop/Phenology Publication/Phenology_R_Scripts")
archippus <- read.csv("L.archippus_observations_1-23-2024.csv")
head(archippus)

length(archippus$id)
a <- subset(archippus, select = c(id, observed_on, latitude, longitude, scientific_name) )
head(a)

archippus <- na.omit(a)
length(archippus$id)

# Adding dates to whole archippus dataset
a.dates <- parse_date_time(archippus$observed_on, "ymd")
a.months <- months(a.dates)
a.year <- year(a.dates)
a.day <- day(a.dates)
days_of_the_year <- yday(a.dates) #  days since Jan 1
archippus$julian_day <- days_of_the_year
archippus$year <- a.year
archippus$month <- a.months
archippus$day <- a.day
archippus

#summary(archippus)
archippus
```

# create latitudinal bins, subset data by years 

```{r B. philenor}
# we are only using data from 2019-2023, when iNaturalist data density increased significantly 
battus <- subset(battus, year %in% 2019:2023)


# Next use the round() function to create a new column binning all latitudes into integer values.
battus$latitudinal_bin <- round(battus$latitude)
battus <- battus[order(battus$latitudinal_bin, decreasing=F),]
battus

# Check to ensure the data was binned correctly.
head(battus)
tail(battus) 
length(battus$id) 
unique(battus$year)
```

```{r L. a. astyanax}
# First subset all data by the total latitudinal range of interest.
astyanax <- subset(astyanax, year %in% 2019:2023)

# Next use the round() function to create a new column binning all latitudes into integer values.
astyanax$latitudinal_bin <- round(astyanax$latitude)
astyanax <- astyanax[order(astyanax$latitudinal_bin, decreasing=F),]
astyanax

# Check to ensure the data was binned correctly.
head(astyanax)
tail(astyanax) 
length(astyanax$id) 
unique(astyanax$year)
```

```{r L. archippus }
# First subset all data by the total latitudinal range of interest.
archippus  <- subset(archippus, year %in% 2019:2023)

# Next use the round() function to create a new column binning all latitudes into integer values.
archippus$latitudinal_bin <- round(archippus$latitude)
archippus <- archippus[order(archippus$latitudinal_bin, decreasing=F),]
archippus

# Check to ensure the data was binned correctly.
head(archippus)
tail(archippus) 
length(archippus$id)
unique(archippus$year)
```

#cap sightings by latitude and date 
this helps correct for artificial inflation of sightings on specific dates (for example, if there was a butterfly surveying event that caused an abnormally high number of butterflies to be recorded in iNaturalist)

```{r all species}
battus$lat_obs <- paste(battus$latitudinal_bin, battus$observed_on, sep= ":")
astyanax$lat_obs <- paste(astyanax$latitudinal_bin, astyanax$observed_on, sep= ":")
archippus$lat_obs <- paste(archippus$latitudinal_bin, archippus$observed_on, sep= ":")

battus_capped = battus[!duplicated(battus$lat_obs),]
astyanax_capped = astyanax[!duplicated(astyanax$lat_obs),]
archippus_capped = archippus[!duplicated(archippus$lat_obs),]
```

# calculate sample size 
calculate total number of observations per species from iNaturalist within the latitudinal bounds of the study (western limit of -98, northern limit of 41, which corresponds to the point at which battus becomes very rare, according to previous research (Ries&Mullen2008))

```{r subsetting by lat and long limits}
battus1 <- subset(battus, longitude >-98)
battus2 <- subset(battus1, latitudinal_bin %in% 30:41)

astyanax1 <- subset(astyanax, longitude >-98)
astyanax2 <- subset(astyanax1, latitudinal_bin %in% 30:41)

archippus1 <- subset(archippus, longitude >-98)
archippus2 <- subset(archippus1, latitudinal_bin %in% 30:41)

```

```{r re-cap data}
battus_capped2 = battus2[!duplicated(battus2$lat_obs),]
astyanax_capped2 = astyanax2[!duplicated(astyanax2$lat_obs),]
archippus_capped2 = archippus2[!duplicated(archippus2$lat_obs),]
bat_total <- length(battus_capped2$id)
ast_total <- length(astyanax_capped2$id)
arc_total <- length(archippus_capped2$id)

all_species_total <- bat_total + ast_total + arc_total
```

# point plots 
```{r create sf options for plotting}

## point plots (not for kernel density)
bat1 <- st_as_sf(battus_capped, coords = c("longitude", "latitude"))
b1 <- ggplot() + geom_sf(data = bat1, color = "lightpink", alpha = 0.5) + theme_bw()

ast1 <- st_as_sf(astyanax_capped, coords = c("longitude", "latitude"))
as1 <- ggplot() + geom_sf(data = ast1, color = "royalblue", alpha = 0.5) + theme_bw()

arc1 <- st_as_sf(archippus_capped, coords = c("longitude", "latitude"))
ar1 <- ggplot() + geom_sf(data = arc1, color = "olivedrab", alpha = 0.5) + theme_bw()

#name world map for point plots below 
world <- map_data("world")
```

```{r battus point plot}
bat_pp <- ggplot() + 
  geom_polygon( data=world, aes(x=long, y=lat, group=group), color="grey", fill = "white", alpha=0.7) + 
  labs(x = "Longitude", y = "Latitude")  +
  geom_sf(data = bat1, color = "navyblue", alpha = 0.3) +
  geom_rect(aes(xmin = -98, xmax = -74, ymin = 29.5, ymax = 41.5),
            fill = "transparent", color = "black", size = 2) +
  coord_sf(xlim = c(-130, -60), ylim = c(15, 60), expand = FALSE) +
  annotate("text", x=-122, y=56, label= "n = 7515", size = 6) + 
  theme_classic() + 
  theme(axis.ticks = element_line(size = 1, color="black") , 
        axis.ticks.length = unit(.3, "cm"), 
        axis.title = element_text( color="black", size=15, face=2),
        axis.line = element_line(size = 1, colour = "black", linetype=1),
        axis.text = element_text(color="black", size=15, face=1)
  ) 
bat_pp
```
```{r astyanax point plot}
ast_pp <- ggplot() + 
  geom_polygon( data=world, aes(x=long, y=lat, group=group), color="grey", fill = "white", alpha=0.7) + 
  labs(x = "Longitude", y = "Latitude")  +
  geom_sf(data = ast1, color = "lightblue", alpha = 0.3) +
  geom_rect(aes(xmin = -98, xmax = -74, ymin = 29.5, ymax = 41.5),
            fill = "transparent", color = "black", size = 2) +
  coord_sf(xlim = c(-130, -60), ylim = c(15, 60), expand = FALSE) +
  annotate("text", x=-122, y=56, label= "n = 6618", size = 6) + 
  theme_classic() + 
  theme(axis.ticks = element_line(size = 1, color="black") , 
        axis.ticks.length = unit(.3, "cm"), 
        axis.title = element_text( color="black", size=15, face=2),
        axis.line = element_line(size = 1, colour = "black", linetype=1),
        axis.text = element_text(color="black", size=15, face=1)
  ) 
ast_pp
```

```{r}
arc_pp <- ggplot() + 
  geom_polygon( data=world, aes(x=long, y=lat, group=group), color="grey", fill = "white", alpha=0.7) + 
  labs(x = "Longitude", y = "Latitude")  +
  geom_sf(data = arc1, color = "orange", alpha = 0.3) +
  geom_rect(aes(xmin = -98, xmax = -74, ymin = 29.5, ymax = 41.5),
            fill = "transparent", color = "black", size = 2) +
  coord_sf(xlim = c(-130, -60), ylim = c(15, 60), expand = FALSE) +
  annotate("text", x=-122, y=56, label= "n = 4805", size = 6) + 
  theme_classic() + 
  theme(axis.ticks = element_line(size = 1, color="black") , 
        axis.ticks.length = unit(.3, "cm"), 
        axis.title = element_text( color="black", size=15, face=2),
        axis.line = element_line(size = 1, colour = "black", linetype=1),
        axis.text = element_text(color="black", size=15, face=1)
  ) 
arc_pp
```
```{r all species figure}
all_species <- ggplot() + 
  geom_polygon( data=world, aes(x=long, y=lat, group=group), color="grey", fill = "white", alpha=0.7) + 
  labs(x = "Longitude", y = "Latitude")  +
  geom_sf(data = arc1, color = "orange", alpha = 0.3) +
  geom_sf(data = ast1, color = "lightblue", alpha = 0.3) +
  geom_sf(data = bat1, color = "navyblue", alpha = 0.3) +
  geom_rect(aes(xmin = -98, xmax = -74, ymin = 29.5, ymax = 41.5),
            fill = "transparent", color = "black", size = 2) +
  coord_sf(xlim = c(-130, -60), ylim = c(15, 60), expand = FALSE) +
  annotate("text", x=-122, y=56, label= "n = 18938", size = 6) + 
  theme_classic() + 
  theme(axis.ticks = element_line(size = 1, color="black") , 
        axis.ticks.length = unit(.3, "cm"), 
        axis.title = element_text( color="black", size=15, face=2),
        axis.line = element_line(size = 1, colour = "black", linetype=1),
        axis.text = element_text(color="black", size=15, face=1)
  ) 

all_species

library(patchwork)

final_fig <- bat_pp + ast_pp + arc_pp + all_species

ggsave(final_fig, file = "/Users/abbyrobinson/Desktop/Phenology Publication/phenology_analysis/final_point_plot.pdf", width=15, height=10, units=c("in"), useDingbats=FALSE)
```

